---
title: "Workshop"
author: "Awtum"
date: "July 12, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


load libraries and datasets
#source("http://bioconductor.org/biocLite.R")
#biocLite("Biobase")
#biocLite("genefilter")
#devtools::install_github("genomicsclass/GSE5859")
```{r library load}
library(Biobase)
library(rafalib)
library(genefilter)
library(GSE5859)
data(GSE5859)
```
## R markdown report 1

### gene expression levels  
```{r}
geneExpression = exprs(e)

```
print date info
```{r}
sampleInfo = pData(e)
head(sampleInfo$date)
```
Output: print a table of year vs ethnic groups
```{r}
year = factor( format(sampleInfo$date,"%y") )
tab = table(year,sampleInfo$ethnicity)
print(tab)
```
###Differential expression analysis
remove control genes
```{r}
out <- grep("AFFX",rownames(geneExpression))
```

Use a t-test to perform differential expression analysis 
between CEU and ASN
```{r}
eth <- sampleInfo$ethnicity
ind<- which(eth%in%c("CEU","ASN"))
res1 <- rowttests(geneExpression[-out,ind],droplevels(eth[ind]))
```
Use a t-test to perform differential expression analysis between 2002 vs 2003 within CEU
```{r}
ind <- which(year%in%c("02","03") & eth=="CEU")
res2 <- rowttests(geneExpression[-out,ind],droplevels(year[ind]))
```
output: plot histograms for p-values 
```{r}
hist(res1$p, main="p-value of CEU vs ASN")
hist(res2$p, main="p-value of 2002 vs 2003 for CEU")
```

## R markdown report 2
### remove a redundant sample
```{r}
cors <- cor(exprs(e))
Pairs=which(abs(cors)>0.9999,arr.ind=TRUE)
out = Pairs[which(Pairs[,1]<Pairs[,2]),,drop=FALSE]
if(length(out[,2])>0) e=e[,-out[2]]
```
remove control probes from the analysis
```{r}
out <- grep("AFFX",featureNames(e))
e <- e[-out,]
```
Create a detrended gene expression data matrix and 
extract the information on dates and ethnic group 
```{r}
y <- exprs(e)-rowMeans(exprs(e))
dates <- pData(e)$date
eth <- pData(e)$ethnicity
```

perform PCA
```{r}
s <- svd(y)
```
output: plot MDS plot of PC1 and PC2 for ethnic groups
```{r}
cols = as.numeric(eth)
mypar()
plot(s$v[,1],s$v[,2],col=cols,pch=16, xlab="PC1",ylab="PC2")
legend("bottomleft",levels(eth),col=seq(along=levels(eth)),pch=16)
title("Ethnic groups")
```
output: plot PC1 and PC2 for years
```{r}
cols = as.numeric(year)
mypar()
plot(s$v[,1],s$v[,2],col=cols,pch=16, xlab="PC1",ylab="PC2")
legend("bottomleft",levels(year),col=seq(along=levels(year)),pch=16)
title("Years")
```
